---
title: 'Determinants of a Valuable Player: Analyzing Key Factors for Player and Team
  Success'
date: "2025-12-05" 
author: "Darrell Liu, Alejandro Leon, Kelvin Kim"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)

# Load cleaned dataset
load(here::here("data", "nba2024_cleaned.RData"))

# Load required libraries
source(here::here("00_requirements.R"))
```


# Abstract 

  NBA player performance is influenced by a wide range of statistical, physical, and gameplay factors. The purpose of this study is to identify the most significant predictors of scoring performance and overall player efficiency during the 2024 NBA season, focusing on key factors such as steals, blocks, minutes played, points scored, shooting percentages, and other advanced efficiency metrics. Using data from nba2024.csv, we apply exploratory analysis and preliminary statistics modeling in order to examine how these variables relate to player performance across the NBA.

  Our results indicate that efficiency metrics such as shooting percentage and usage rate are the strongest predictors of scoring performance, showing a clear positive correlation. Traditional statistics such as rebounds and assists provide additional context but demonstrate weaker associations. Several variables such as scoring and steals required adjustments to address skewing and improve the model fit. In addition certain severely outlying player statistics had to be removed in order to prevent skewing of the model. These findings offer valuable insight for analysts, coaches, fans, and players themselves in order to highlight the metrics that most accurately reflect player value and on-court impact. 

Understanding these relationships supports more informed evaluation, prediction, and strategic decision making in the NBA for the future.


# Introduction
  The purpose of this analysis is to explore and better understand the statistical drivers of player performance in the NBA during the 2024 season. Our central question is: “Which player statistics most strongly influence scoring and overall efficiency?”

  This analysis can benefit basketball analysts, coaches, fantasy basketball participants, and anyone interested in data-driven sports insights. While we are not yet training a full predictive model, we are preparing the foundation for statistical modeling by examining variable behavior, relationships, and distribution patterns across the league. Through this project, we aim to identify meaningful performance trends and generate hypotheses for future modeling work, such as predicting high-value players.


# Data
  Our analysis uses the dataset nba2024.csv, which contains full-season statistics for NBA players during the 2024 season. The dataset includes variables such as points per game, rebounds, assists, minutes played, field goal percentage, three-point percentage, and advanced metrics where available. These data were retrieved from publicly available NBA statistics sources and compiled into a cleaned .csv file for the project.

  To prepare the dataset, we removed players with missing or incomplete statistical entries, standardized variable formats, and ensured that numeric fields were correctly encoded. No simulated data were used; the analysis is entirely based on real-world NBA statistics.

# Visualizations

  The visualizations displayed below are of the categories we will be looking for in determining which player statistics are most strongly associated with high scoring performance, overall efficiency, and player value. 

  We want to see how much team performance and statistics change a player’s value, whether the numeric values matter more or the percentage of the team stats matters more, and where do the top players most stand out as outliers and certain statistics. 

  Currently we can see that team statistics do somewhat contribute to a player's success, although there is no direct correlation or causation for the claim. Durability and minutes players have a small impact, as players do need to play a lot of minutes to be highly productive players, but there isn’t a complete linear positive correlation for that claim. We also see that position doesn’t affect a player's value, as all positions have relatively the same average points per game.
  
#Analysis
```{r}
nba <- nba2024
```


```{r}
hist(nba$PTS,
     main = "Distribution of Points Per Game",
     xlab = "PPG",
     col = "lightblue",
     border = "white")
```

This histogram illustrates a right-skewed (positively skewed) distribution of Points Per Game (PPG). The shape indicates that the vast majority of NBA players fall into the lower scoring ranges, while only a small number of players achieve very high scoring averages, creating a long tail to the right. The median, or the most frequent range, is heavily concentrated between approximately 10 and 15 PPG, mostly representing the rotational players in the NBA.

The distribution shows significant concentration in the lower-to-moderate scoring tiers, with the bulk of the data falling below 20 PPG. This high concentration highlights the fundamental structure of an NBA roster, where many players fulfill specialized, lower-scoring roles. However, the presence of bars extending toward 30 PPG represents the elite scorers and superstars, whose exceptionally high averages demonstrate the significant variability in scoring output across the league.



